formula1

### 2020-06-11 inst/f1points.Rmd
more work

Formula 1 racing is an important and prestigious motor sport
\citep{moul2009}.  Season ranking is based on a points system; points
are awarded to drivers and teams based on race order, and accumulate
additively.  The current points system, for example, is
(25-18-15-12-10-8-6-4-2-1) [that is, the winner is awarded 25 points,
second place 18 points, and so on], although since 2019 the fastest
lap attracts one point.

@@

Points systems similar to that of Formula 1 are common in other sports
and examples would include modern pentathlon, curling, chess, rowing,
and figure skating; all have the common feature of interpretation of
an order statistic to a points.  Further, we see points systems
frequently in the wider context of competitive situations such as the
Eurovision Song Contest, Australian MasterChef, University rankings,
educational league tables, and car safety ratings.

@@

@@

We see the Hamilton has the largest estimated strength, although it
should be noted that a likelihood ratio test [`samep.test()`] does not
reject the null of $p_\mathrm{Hamilton}=p_\mathrm{Vettel}$ ($p=0.09$).

```{r,label=dosameptest,cache=TRUE}
samep.test(H,1:2)
```

@@

```{r,label=calcpoints,echo=TRUE,cache=TRUE}
p <- c(25, 18, 15, 12, 10, 8, 6, 4, 2, 1, 0, 0)
points_real <- p
dp <- ordertable2points(as.ordertable(F1_table_2017[seq_len(top),]),p)
dp <- sort(dp,decreasing=TRUE)
dp[] <- seq_along(dp)
m <- maxp(H)
m <- sort(m,decreasing=TRUE)
m[] <- seq_along(m)
```

@@

From the above, we see that the first two drivers (Hamilton, Vettel)
are are respectively first and second according to both ranking
ranking procedures; but Raikkonen and Bottas are third and fourth, and
fourth and third, according to points and likelihood respectively.
The degree of agreement between the two systems might be quantified as
"two", for the positions of the top two drivers agree.

We can use this observation to assess a given points system: using
random sampling techniques we can generate an order table \emph{in
silico}, using estimated driver strengths from the 2017 table.  Given
this table we can then then compare rankings generated by maximum
likelihood \citeauthor{plackett1975} strengths with rankings generated
by any particular points system; this furnishes an objective
assessment of a

```{r,label=montecarlosetup, echo=FALSE, cache=FALSE}
top <- 11
f2017 <- read.table("formula1_2017.txt",header=TRUE)[seq_len(top),1:20]
#m <- maxp(ordertable2supp(as.ordertable(f2017)))

## Now work on synthetic dataset generation:
resampling <- function(m,f2017,points){
    random_table <- rrank(n=ncol(f2017), p=m)
    rownames(random_table) <- colnames(f2017)
    ## "random_table" is a random table; now calculate lstar and pstar for
    ## likelihood ranks and points ranks respectively:

    p_star <- ordertable2points(as.ordertable(ranktable_to_ordertable(random_table)),points)
    p_star <- sort(p_star,decreasing=TRUE)
    p_star[] <- seq_along(p_star)

    goodness <- sum(cumprod(names(p_star)==names(l_star)))
    return(goodness)
}
```

```{r,label=montecarlosetup2, cache=TRUE,echo=FALSE}
f2017 <- read.table("formula1_2017.txt",header=TRUE)[seq_len(top),1:20]
m <- maxp(ordertable2supp(as.ordertable(f2017)))
```

```{r,label=montecarlosetup3,echo=FALSE}
o_real <- function(n){replicate(n,resampling(m,f2017,points=points_real))}
```

```{r,label=montecarlorealpointsdoer,cache=TRUE,echo=FALSE}
or <- o_real(1000)
```

```{r,label=defineotop, cache = TRUE, echo=FALSE}
o_top <- function(top,n=100){replicate(n,resampling(m,f2017,points=rep(1,top)))}
```

```{r,label=executeotop1, cache = TRUE, echo=FALSE}
ot1 <- o_top(1,n=1000)
```

```{r,label=executeotop2, cache = TRUE,echo=FALSE}
ot2 <- o_top(2,n=1000)
```

```{r,label=executeotop3, cache = TRUE,echo=FALSE}
ot3 <- o_top(3,n=1000)
```

```{r,label=executeotop4, cache = TRUE,echo=FALSE}
ot4 <- o_top(4,n=1000)
```

```{r,label=executeotop5, cache = TRUE,echo=FALSE}
ot5 <- o_top(5,n=1000)
```

```{r,label=defineotherschemes, echo=FALSE}
o_zip <- function(n=100){replicate(n,resampling(m,f2017,points=1/seq_len(top)))}
o_exp <- function(p,n=100){replicate(n,resampling(m,f2017,points=p^seq_len(top)))}
```

```{r,label=ozcalculator,cache=TRUE, echo=FALSE}
oz <- o_zip(1000)
```

```{r, label=oecalculator, cache=TRUE, echo=FALSE}
oe <- o_exp(0.5,1000)
```

```{r, label=makeManddimlabel, echo=FALSE}
M <- rbind(
	tabulate(or +1,nbins=12),
	tabulate(ot1+1,nbins=12),
	tabulate(ot2+1,nbins=12),
	tabulate(ot3+1,nbins=12),
	tabulate(ot4+1,nbins=12),
	tabulate(ot5+1,nbins=12),
	tabulate(oz +1,nbins=12),
	tabulate(oe +1,nbins=12)
	)
dimnames(M) <- list(system=c("real",paste("top",seq_len(5),sep=""),"zipf","exp"),correct=0:11)
```

```{r,showM}
M
```

```{r,showcumsumM}
M2 <- t(apply(M[,12:1],1,cumsum))[,12:1]
dimnames(M2) <- list(system=rownames(M),correct=paste(">=",colnames(M2)))
M2
matplot(t(M2[,12:1]),type='b')
```

# more efficient ways

```{r,label=multiplemontecarlosetup, echo=FALSE, cache=FALSE}
top <- 11
f2017 <- read.table("formula1_2017.txt",header=TRUE)[seq_len(top),1:20]
#m <- maxp(ordertable2supp(as.ordertable(f2017)))

@@

```{r,label=multiplemontecarloexecute, cache=TRUE,echo=TRUE}
f2017 <- read.table("formula1_2017.txt",header=TRUE)[seq_len(top),1:20]
m <- maxp(ordertable2supp(as.ordertable(f2017)))
OO <- replicate(500+500,resampling_multiple(m,f2017,pointslist=pointslist))
```

```{r,label=multiplemontecarlotabulation}
M3 <- t(apply(OO,1,function(o){tabulate(o+1,nbins=12)}))
dimnames(M3) <- list(system=names(pointslist),correct=0:11)
M3
Mc <- t(apply(M3,1,cumsum))
Mc
```

and visuial:

@@

>>>>

Formula 1 racing is an important and prestigious motor sport
\citep{moul2009}.  Season ranking is based on a points system; points
are awarded to drivers and teams based on race order, and accumulate
additively.  The current points system, for example, is
(25,18,15,12,10,8,6,4,2,1) [that is, the winner is awarded 25 points,
second place 18 points, and so on], although since 2019 the fastest
lap attracts one point.

@@

Points systems similar to that of Formula 1 are common in other sports
and examples would include modern pentathlon, curling, chess, rowing,
and figure skating; all have the common feature of translating ranks
into scores which combine additively to generate an overall ranking.
Further, we see points systems frequently in the wider context of
competitive situations such as the Eurovision Song Contest, Australian
MasterChef, University rankings, educational league tables, car safety
ratings, and many others.

@@

## Formula 1 racing: points systems

The points system of Formula 1 Motor racing is a contentious topic;
there have been many changes since systems has changed many times,
although some changes involved only bonus points (typically, one point
for fastest lap), which are not considered here.  The inaugural
system, used from 1950 to 1959 was $(8,6,4,3,2)$, thus crediting only
the first five finishers, and the current system of
$(25,18,15,12,10,8,6,4,2,1)$ credits the first 10.  Arguably these two
systems could introduce different rational behaviour under zero-sum
assumptions: if, in a race, a driver knows he will place seventh under
a low-risk strategy but may place sixth by dint of driving more
aggressively, the low-risk strategy might be rational under the first
points system (which does not reward the extra ranking), but not under
the second, which does.  Still, it is reasonable to assume that each
driver strives to maximise his rank.

@@

We see the Hamilton has the largest estimated strength, although it
should be noted that a likelihood ratio test [`samep.test()`, supplied
with the package] does not reject the null of
$p_\mathrm{Hamilton}=p_\mathrm{Vettel}$ , with a likelihood ratio of
$e^{1.5}\simeq 4.48$, corresponding to an asymptotic $p$-value of
about $0.08$.

```{r,label=dosameptest, echo=FALSE, print=FALSE, cache=TRUE}
ignore <- samep.test(H,1:2)
```

@@

```{r, label=calcpoints, echo=FALSE, cache=TRUE}
p <- c(25, 18, 15, 12, 10, 8, 6, 4, 2, 1, 0, 0)
points_real <- p
dp <- ordertable2points(as.ordertable(F1_table_2017[seq_len(top),]),p)
dp <- sort(dp,decreasing=TRUE)
dp[] <- seq_along(dp)
m <- maxp(H)
m <- sort(m,decreasing=TRUE)
m[] <- seq_along(m)
```

@@

From the above, we see that the first two drivers (Hamilton, Vettel)
are are respectively first and second according to both ranking
ranking procedures; but RAikkOnen 
*crap problem: latex ignores  diaeresis directive* and Bottas are third and fourth, and fourth and
third, according to points and likelihood respectively.  The degree of
agreement between the two systems might be quantified as "two", for
the positions of the top two drivers agree.

We can now assess objectively a given points system: using random
sampling techniques we can repeatedly generate an order table \emph{in
silico}, using estimated driver strengths from the 2017 table.  Given
a synthetic table, calculate drivers' maximum likelihood
\citeauthor{plackett1975} strengths, and also their points awarded
according to any given points system.  We then compare rankings
generated by the \citeauthor{plackett1975} strengths and the points
awarded and note the degree of agreement between the two.  This
furnishes an objective assessment of the points system used.

# Numerical results

Here we follow the above protocol to assess different plausible points
systems.  There are a number of plausible points systems that might be
used:

\begin{itemize}
\item winner takes all $(1,0,0,\ldots)$
\item Zipfs law $(1,1/2,1/3,1/4,\ldots)$
\item exponential decay $(1,r,r^2,r^3,\ldots)$
\item the actual real one $(25,18,15,12,10,8,6,4,2,1)$
\item equal points to the top $n$ competitors $\underbrace{1,1,\ldots
,1}_{n},0,0,\ldots$ (noting that $n=1$ corresponds to winner takes all)
\item linear decrease $(n,n-1,n-2,\ldots,3,2,1,0)$
\end{itemize}

```{r,label=definemultiplemontecarlo, echo=FALSE}
top <- 11
f2017 <- read.table("formula1_2017.txt",header=TRUE)[seq_len(top),1:20]
m <- maxp(ordertable2supp(as.ordertable(f2017)))

@@

```{r,label=multiplemontecarloexecute, cache=TRUE,echo=FALSE}
f2017 <- read.table("formula1_2017.txt",header=TRUE)[seq_len(top),1:20]
m <- maxp(ordertable2supp(as.ordertable(f2017)))
OO <- replicate(500+500,resampling_multiple(m,f2017,pointslist=pointslist))
```

```{r,label=multiplemontecarlotabulation,echo=FALSE}
M3 <- t(apply(OO,1,function(o){tabulate(o+1,nbins=12)}))
dimnames(M3) <- list(system=names(pointslist),correct=0:11)
M3
Mc <- t(apply(M3,1,cumsum))
Mc
```

and visual:

@@

```{r,makeatable}
table(real_winner_ok=OO[1,]>0,wta_winner_ok=OO[3,]>0)
table(real_all=OO[1,]==11,wta_all=OO[3,]==11)
write.csv(OO,file="~/f.txt")
```

# Discussion and conclusions
